Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-03-12 draft.tex
added more figures

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to all decay or scattering
processes that can be formulated in terms of an orthogonal basis of angular functions.
We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. \add{As an example,
        see the LHCb analysis of \dots}.
        The method of moments does not require reparametrizations or approximations.

    \item \add{Likelihood fits  can be unstable in case the underlying
        physical model is only partially known. This can lead to
        overestimating the number of physical parameters, and consequently
        inhibits the convergence of the fits. As an example, this type
        of problem occured in toy studies of the decay $B\to K^*(\to K\pi)\ell^+\ell^-$
        as reported in \cite{Egede:1142152}. It was subsequently solved
        when a missing symmetry relation between the angular observables
        was found and applied, thereby reducing the number of fit parameters.
    }
    \item Mismodeling the underlying physics model can result in systematic bias
        in likelihood fits. We will
        show that the method of moments is insensitive to a certain type of mismodeling;
        i.e. introducing a cutoff in a partial-wave expansion of the signal PDF.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ as
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
where $\lbrace \dual{f}_i \rbrace$ denotes a \emph{dual basis} of angular functions
\add{, and $\Omega$ represents the entire angular phase space.}. In general, $\lbrace \dual{f}_i \rbrace$ may differ from $\lbrace f_i \rbrace$. This is the
case for our selection of applications in appendices \ref{app:btokll} through \ref{app:btokstarll}.\\

@@

    \item We also compare the results as obtained by the method of moments with
        results obtained by a conventional likelihood fit. In particular, we
        study the correlation between the method-of-moment estimators and the
        maximum-likelihood-fit estimators. We run $10^3$ toy analyses, with
        $200$ simulated events per analysis. We show the joint distribution of
        the two estimators in \reffig{correlation}. The two estimators are
        highly correlated. The distribution of the difference of the estimators
        exhibits now bias, which is due to the large number of simulated
        events.  Still, we find that statistical uncertainty on the difference
        of the two estimators is sizeable and can easily become half as large
        as the statistical uncertainty of either estimator.
\end{itemize}
\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S4.png}
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S8.png}
        \caption{Uncertainty of the angular observables $S_4$ (uppwe) and $S_8$
            (lower), extracted from $2\cdot 10^5$ studies of simulated events
            for the decay $B\to K^*\ell^+\ell^-$. We show the uncertainty as a
            function of the number of simulated events $N$.  The red curve
        represents a fit to the function given in \refeq{unc-on-mean}.}
        \label{fig:errors}
\end{figure}

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/S3_scat.pdf}
            \includegraphics[width=0.45\textwidth]{figs/S5_scat.pdf}
            \caption{The joint distribution of the two estimators
                $\est{S}_i^\text{MoM}$ and $\est{S}_i^\text{Fit}$ that arise
                from the method of moments and a maximum-likelihood fit,
                respectively. We show the estimators for both of our benchmark
                observables $S_3$ (upper) and $S_5$ (lower).
                \danny{Marcin: please replace by $S_5$ (upper) and $S_7$ (lower).}
            }
        \label{fig:correlation}
\end{figure}

>>>>

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to all decay or scattering
processes that can be formulated in terms of an orthogonal basis of angular functions.
We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. As an example see LHCb measurement of $\bar{B}\to \bar{K}^*(\to \bar{K}\pi)\mu^+\mu^-$\cite{Aaij:2013qta}. The method of moments does not require reparametrizations or approximations.

    \item Likelihood fits  can be unstable in case the underlying
        physical model is only partially known. This can lead to
        overestimating the number of physical parameters, and consequently
        inhibits the convergence of the fits. As an example, this type
        of problem occured in toy studies of the decay $B\to K^*(\to K\pi)\ell^+\ell^-$
        as reported in \cite{Egede:1142152}. It was subsequently solved
        when a missing symmetry relation between the angular observables
        was found and applied, thereby reducing the number of fit parameters.
    \item Mismodeling the underlying physics model can result in systematic bias
        in likelihood fits. We will
        show that the method of moments is insensitive to a certain type of mismodeling;
        i.e. introducing a cutoff in a partial-wave expansion of the signal PDF.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ as
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
where $\lbrace \dual{f}_i \rbrace$ denotes a \emph{dual basis} of angular functions, and $\Omega$ represents the entire angular phase space. In general, $\lbrace \dual{f}_i \rbrace$ may differ from $\lbrace f_i \rbrace$. This is the
case for our selection of applications in appendices \ref{app:btokll} through \ref{app:btokstarll}.\\

@@

    \item We also compare the results as obtained by the method of moments with
        results obtained by a conventional likelihood fit. In particular, we
        study the correlation between the method-of-moment estimators and the
        maximum-likelihood-fit estimators. We run $10^3$ toy analyses, with
        $200$ simulated events per analysis. We show the joint distribution of
        the two estimators in \reffig{correlation}. The two estimators are
        highly correlated. The distribution of the difference of the estimators
        exhibits now bias, which is due to the large number of simulated
        events.  Still, we find that statistical uncertainty on the difference
        of the two estimators is sizeable and can easily become half as large
        as the statistical uncertainty of either estimator.
\end{itemize}
\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S5.pdf}
            \includegraphics[width=0.45\textwidth]{figs/Q2_1_2_S9.pdf}
        \caption{Uncertainty of the angular observables $S_5$ (upper) and $S_9$
            (lower), extracted from $2\cdot 10^5$ studies of simulated events
            for the decay $B\to K^*\ell^+\ell^-$. We show the uncertainty as a
            function of the number of simulated events $N$.  The red curve
        represents a fit to the function given in \refeq{unc-on-mean}. Errors correspond to $68~\%$ spread of measured uncertainties in the toys.}
        \label{fig:errors}
\end{figure}

\begin{figure}[t]
        \centering
            \includegraphics[width=0.45\textwidth]{figs/S5_scat.pdf}
            \includegraphics[width=0.45\textwidth]{figs/S7_scat.pdf}
            \caption{The joint distribution of the two estimators
                $\est{S}_i^\text{MoM}$ and $\est{S}_i^\text{Fit}$ that arise
                from the method of moments and a maximum-likelihood fit,
                respectively. We show the estimators for both of our benchmark
                observables $S_3$ (upper) and $S_5$ (lower).
                \danny{Marcin: please replace by $S_5$ (upper) and $S_7$ (lower).}
            }
        \label{fig:correlation}
\end{figure}
